Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-27 paper/paper.tex

%%%%%%%%%%%%%%%%%%%%%%

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================
In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
 and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.

%\comment{

\section{Generating Liveness Equations}
\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

 For our example program which has  a
single function \length, this generates the following equations:

  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \epsilonset \cup \acdr\Lf{\length}{1}{\epsilon}
   \end{eqnarray*}

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
\end{align*}

>>>>

\begin{figure*}[t!]
%%\begin{picture}(100,130)(0,-90)
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF$_{ep_3}$\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_{ep_1}~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_{ep_2}~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN~$\pi_{11}\!\!:\,$ (\RETURN$_{ep_4}$\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
%% >>>>>>> 67d5a8ebe45703d9726af91b752e9d74152ae455
\end{figure*}

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \pu\ is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.

In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.
\section{Computing liveness and its encoding as a table}\label{sec:computing}

Section~\ref{sec:liveness} gave a context-sensitive liveness
analysis and {\color {red}  proved it correct with reference
  to   a  {\em   minefield}   semantics}.   In   particular,
\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
describes  a  liveness  set  for  each program  point  in  a
function in terms of a  given $\sigma$ and \Lfonly. We still
have not mentioned what is the specific sigma that should be
used for each function How is LF to be computed.

It helps to cast the equations  in the form of a grammar. To
do  this  we  first  have  to  cast  massage  the  equations
slightly:

\subsubsection{Modifying the liveness rules.}

The      $\mathit{ref}$     rule      for     \CONS,      shown     in
Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
$\emptyset$, if $\sigma$ itself $\emptyset$. This creates difficulties
since  we  want  to  solve  the  equations  arising  out  of  liveness
symbolically, and realizing these rules requires $\sigma$ to be known.

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}

and \cbar\ as:

\begin{align*}
 \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}

We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,l \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.

The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \bcdr\sigma_2',~\mbox{and}\\  
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

\begin{proposition}
Assume that a liveness computation  based on the original set of rules
gives the liveness  of the variable $x$ at a  program point $\pi_i$ as
$\sigma$   (symbolically,   $\Lanv{i}{x}=   \sigma$).   Further,   let
$\Lanv{i}{x}=  \sigma'$ when the  modified rules  are used  instead of
\Lfunonly.  Then $\sigma' \stackrel{*}{\hookrightarrow} \sigma$.
\end{proposition}

For  an explanation  of why  the  proposition holds  for the  modified
\CONS\  rule,   we  refer   the  reader  to   \cite{asati14lgc}.   The
proposition also holds for the the other rules for similar reasons.

%% To see why the  proposition is true, consider an analysis
%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
%% marks a place in  $\alpha$ where the original \CONS\ rule
%% would have erased  an immediately following \acar(\acdr),
%% or,  in absence  of  such a  symbol,  would have  dropped
%% $\alpha$ itself.   Since the  application of any  rule in
%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
%% $\alpha$, the  markers and other symbols  in $\alpha$ are
%% propagated to  other dependent parts of  program in their
%% same  relative   positions.   Consequently,  the  erasure
%% carried   out   at  the   end   of   the  analysis   with
%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
%% obtained through \Lfunonly.

\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}

\section{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
how  to generate  the equations  describing the  demand transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine    \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   Of   particular
interest  is the liveness  at the  program point  at the  beginning of
beginning  of $e_{\mathit{f}}$  (assumed to  be $\pi_1$).  We generate
equations  of the  form $\Lf{\mathit{f}}{i}{\sigma}  = \Lanv{1}{x_i}$,
where  $x_i$   is  the  $i^{\mbox{\footnotesize   th}}$  parameter  of
$\mathit{f}$. Doing this for  \length, we get:
\begin{align*}
 \Lanv{1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
\end{align*}
Which gives the following equation for \Lfone{\length}: 
  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
   \end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of \length,  \Lan{1}\ may,  in turn,  be expressed  in  terms of
\Lfone{\mathit{f}}.
